---
title: "tpaHH: Two-Piece Asymmetric Hybrid Hazard Regression Models"
subtitle: "Annotated Source Code and User Manual"
author: "Prepared for Ewnetu-github/tpaHH (originally `Topic3_sourcecode.R`)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    theme: readable
    highlight: tango
    code_folding: hide
    code_download: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center", fig.width = 6, fig.height = 4)
```

# Introduction

This document is a fully commented, reorganised rewrite of the original
`Topic3_sourcecode.R` file from the GitHub repository
[`Ewnetu-github/tpaHH`](https://github.com/Ewnetu-github/tpaHH). The
underlying statistical logic of every function is **unchanged** from the
original script -- only the presentation has been rewritten: functions are
grouped by purpose, every function is documented with a `roxygen2`-style
header (`@param`, `@return`, `@details`), inline comments explain non-obvious
steps, and a worked example at the end shows the full workflow end to end.

## What the toolkit does

The code implements **hybrid hazard-based survival regression models** built
from **two-piece asymmetric (TPA)** baseline distributions. In these models,
a covariate can act on the survival time on two different scales at once:

* a **time-scale** (acceleration) effect, which stretches or compresses time
  itself (as in an Accelerated Failure Time, AFT, model), and
* a **hazard-scale** (relative-risk) effect, which multiplies the hazard up
  or down (as in a Proportional Hazards, PH, model).

Combining both gives the **Hybrid Hazard (HH)** model; letting only one of
the two operate recovers the classical **AFT**, **PH**, **Accelerated
Hazards (AH)**, or **General Hazards (GH)** models as special cases. The
covariate effect on either scale can be modelled:

* **parametrically** (a linear index, e.g. $x^\top\beta$),
* **semi-parametrically** (part linear, part smooth), or
* **non-parametrically** (a fully smooth function estimated by local
  likelihood with a kernel).

Three baseline (error) distributions are supported for the two-piece family:

| Code label   | Baseline distribution   |
|--------------|--------------------------|
| `tpalnorm`   | TPA (log-)normal         |
| `tpallogis`  | TPA (log-)logistic       |
| `tpalLap`    | TPA (log-)Laplace        |

All three share the **same** log-link time transform $g(t)=\log(t)$, so the
resulting failure-time distributions are, respectively, two-piece
log-normal, log-logistic and log-Laplace.

## How this document is organised

1. **Package setup** — the R packages the code depends on.
2. **Baseline distributions and link functions** — the building blocks.
3. **The TPA distribution family** — density, CDF, hazard, cumulative
   hazard, quantile function and random-number generator.
4. **Fully parametric regression models** — `HH`, `GH`, `AFT`, `PH`, `AH`.
5. **Kernel function and semi-/non-parametric estimation** — local-likelihood
   counterparts of the models above.
6. **Post-estimation tools** — prediction, distance measures, numerical
   integration, and confidence intervals.
7. **Worked example** — simulate data, fit a model, obtain standard errors,
   and plot a predicted survival curve.
8. **Function reference table** — a one-page lookup of every function.
9. **Notes on the original code** — a short, honest list of quirks/limits
   inherited from the original script, so users know what to check before
   relying on a given function.

## Requirements

```{r install-note, eval=FALSE}
# Run once, interactively, before using this document.
# CRAN packages:
install.packages(c("nlme", "pracma"))

# Packages not currently on CRAN (or only as an archived version) --
# install from their GitHub mirrors instead:
install.packages("remotes")
remotes::install_github("cran/Deriv")       # symbolic differentiation
remotes::install_github("cran/GoFKernel")   # numerical inversion of a CDF

# QBAsyDist is only used in the original script for its expit() helper
# (the standard logistic/sigmoid function). This document defines expit()
# directly (see Section 2) so QBAsyDist is NOT required to run anything
# here. If you want the full QBAsyDist package for other purposes, see
# https://github.com/cran/QBAsyDist (it depends on several packages that
# are archived on CRAN, so plan for extra install steps).
```

```{r libraries}
suppressPackageStartupMessages({
  library(Deriv)      # symbolic differentiation, used by dtpa()
  library(GoFKernel)  # inverse() : numerically inverts a CDF, used by qtpa()
  library(pracma)     # trapz(): trapezoidal numerical integration
  library(nlme)        # fdHess(): finite-difference Hessian, used for CIs
})

# expit() is the inverse-logit (standard logistic) function.
# The original script pulled this from the QBAsyDist package; because
# QBAsyDist is hard to install (see the note above) we define it directly.
# It is mathematically identical to QBAsyDist::expit().
expit <- function(x) 1 / (1 + exp(-x))
```

\newpage

# Baseline distributions and link functions

Every TPA model is built by "gluing" two halves of a *reference* symmetric
distribution together at a location parameter. The functions below provide
the density (`f0_*`), CDF (`F0_*`) and quantile function (`QF_*`) of each of
the three reference distributions used as building blocks, plus the link
function `g()` that transforms time before the two-piece construction is
applied.

```{r baseline-normal}
#' Standard normal reference distribution
#' @param s numeric vector of evaluation points
#' @return f0_N: density; F0_N: CDF; QF_N: quantile function
#' @export
f0_N <- function(s) dnorm(s, mean = 0, sd = 1)                 # density
F0_N <- function(s) pnorm(s, mean = 0, sd = 1)                 # CDF
QF_N <- function(tau) qnorm(tau, mean = 0, sd = 1,              # quantile
                             lower.tail = TRUE, log.p = FALSE)
```

```{r baseline-laplace}
#' Standard Laplace (double-exponential) reference distribution
#' @param s numeric vector of evaluation points
#' @param beta probability(-ies) in (0, 1) for the quantile function
#' @return f0_La: density; F0_La: CDF; QF_La: quantile function
#' @export
f0_La <- function(s) 0.5 * exp(-abs(s))                                  # density
F0_La <- function(s) 0.5 + 0.5 * sign(s) * (1 - exp(-abs(s)))            # CDF
QF_La <- function(beta) -sign(beta - 0.5) * log(1 - 2 * abs(beta - 0.5)) # quantile
```

```{r baseline-logistic}
#' Standard logistic reference distribution
#' @param s numeric vector of evaluation points
#' @param tau probability(-ies) in (0, 1) for the quantile function
#' @return f0_Lo: density; F0_Lo: CDF; QF_Lo: quantile function
#' @export
f0_Lo <- function(s) exp(s) * (1 + exp(s))^-2   # density
F0_Lo <- function(s) 1 / (1 + exp(-s))          # CDF
QF_Lo <- function(tau) log(tau / (1 - tau))     # quantile (logit)
```

```{r link-functions}
#' Link functions used to transform time before applying the two-piece
#' construction. Only the log-link (glog) is exercised by the parametric
#' and non-parametric estimators in this file; glogit/glogit.inv are
#' provided for completeness / alternative link choices.
#' @param s numeric vector
#' @export
glog       <- function(s) log(s)          # time -> log(time)
glog.inv   <- function(s) exp(s)          # inverse of glog
glogit     <- function(s) log(exp(s) - 1) # alternative link
glogit.inv <- function(s) log(exp(s) + 1) # inverse of glogit
```

\newpage

# The TPA distribution family

These six functions define the two-piece asymmetric (TPA) distribution
built from a symmetric reference distribution (`f0`, `F0`, `QF`) and a link
function `g`. The construction follows Rubio & Steel-type two-piece
distributions: a single location parameter `eta` splits the real line into
two halves, each half is a stretched/compressed copy of the reference
distribution, and `alpha` controls how much probability mass sits in each
half (`alpha = 0.5` recovers the symmetric case).

```{r tpa-density}
#' Density of the TPA family
#'
#' @param t vector of quantiles (time)
#' @param eta location parameter
#' @param phi scale parameter (phi > 0)
#' @param alpha skewness/index parameter in (0, 1)
#' @param f0 density of the reference distribution, e.g. f0_N
#' @param g link function applied to time, e.g. glog
#' @param log logical; if TRUE, the log-density is returned
#' @return numeric vector of (log-)density values, same length as t
#' @export
dtpa <- function(t, eta, phi, alpha, f0, g, log = FALSE) {
  # g.prime is the derivative of the link function, obtained symbolically
  # (this is why the Deriv package is required). For g = glog = log(t),
  # g.prime(t) = 1/t.
  g.prime <- Deriv::Deriv(g)

  # Common multiplicative term shared by both pieces of the density.
  com <- 2 * alpha * (1 - alpha) * g.prime(t) / phi

  # Left piece (t < eta) uses (1 - alpha) as the scale factor;
  # right piece (t >= eta) uses alpha. This is what makes the
  # distribution asymmetric around eta.
  den <- ifelse(t < eta,
                com * f0((1 - alpha) * (g(eta) - g(t)) / phi),
                com * f0(alpha * (g(t) - g(eta)) / phi))

  if (log) return(log(den)) else return(den)
}
```

```{r tpa-cdf}
#' Cumulative distribution function (CDF) of the TPA family
#' @param q vector of quantiles (time)
#' @param eta,phi,alpha TPA parameters (see dtpa)
#' @param F0 CDF of the reference distribution, e.g. F0_N
#' @param g link function, e.g. glog
#' @param lower.tail logical; if FALSE, 1 - CDF (the survival function) is returned
#' @return numeric vector of (survival) probabilities
#' @export
ptpa <- function(q, eta, phi, alpha, F0, g, lower.tail = TRUE) {
  p <- ifelse(q < eta,
              2 * alpha * F0((1 - alpha) * (g(q) - g(eta)) / phi),
              2 * alpha - 1 + 2 * (1 - alpha) * F0(alpha * (g(q) - g(eta)) / phi))
  if (lower.tail) return(p) else return(1 - p)
}
```

```{r tpa-hazard}
#' Hazard function of the TPA family: h(t) = f(t) / S(t)
#' @param t vector of quantiles (time)
#' @param eta,phi,alpha TPA parameters
#' @param f0 reference density
#' @param F0 reference CDF
#' @param g link function
#' @param log logical; if TRUE, the log-hazard is returned
#' @export
htpa <- function(t, eta, phi, alpha, f0, F0, g, log = FALSE) {
  dens <- dtpa(t, eta, phi, alpha, f0, g, log = FALSE)
  surv <- ptpa(t, eta, phi, alpha, F0, g, lower.tail = FALSE)
  h <- dens / surv
  if (log) return(log(h)) else return(h)
}
```

```{r tpa-cumhazard}
#' Cumulative hazard function of the TPA family: H(t) = -log(S(t))
#' @param t vector of quantiles (time)
#' @param eta,phi,alpha TPA parameters
#' @param F0 reference CDF
#' @param g link function
#' @param log logical; if TRUE, log(H(t)) is returned
#' @export
Htpa <- function(t, eta, phi, alpha, F0, g, log = FALSE) {
  cmh <- -log(ptpa(t, eta, phi, alpha, F0, g, lower.tail = FALSE))
  if (log) return(log(cmh)) else return(cmh)
}
```

```{r tpa-quantile}
#' Quantile function of the TPA family
#'
#' @param tau vector of probabilities in (0, 1)
#' @param eta,phi,alpha TPA parameters
#' @param F0 reference CDF
#' @param g link function
#' @param g.inv inverse link function; if NULL it is obtained numerically
#'   via GoFKernel::inverse()
#' @param QF analytic quantile function of the reference distribution
#'   (e.g. QF_N); if NULL it is obtained numerically via GoFKernel::inverse()
#' @return numeric vector of quantiles, same length as tau
#' @export
qtpa <- function(tau, eta, phi, alpha, F0, g, g.inv = NULL, QF = NULL) {
  # Fall back to numerical inversion when analytic forms are not supplied.
  if (is.null(QF)) {
    QF <- GoFKernel::inverse(F0, lower = -Inf, upper = Inf)
  }
  if (is.null(g.inv)) {
    g.inv <- GoFKernel::inverse(g, lower = 0, upper = Inf)
  }

  q <- vector()
  for (i in 1:length(tau)) {
    # Ca is the standardised quantile on the reference scale, computed
    # piecewise depending on which side of alpha the target probability
    # falls on.
    Ca <- ifelse(tau[i] < alpha,
                 (1 / (1 - alpha)) * QF(tau[i] / (2 * alpha)),
                 (1 / alpha) * QF((1 + tau[i] - 2 * alpha) / (2 * (1 - alpha))))
    q[i] <- g.inv(g(eta) + phi * Ca)
  }
  return(q)
}
```

```{r tpa-random}
#' Random number generation from the TPA family (inverse-CDF method)
#' @param n number of draws
#' @param eta,phi,alpha TPA parameters
#' @param F0 reference CDF
#' @param g link function
#' @param QF quantile function of the reference distribution
#' @return numeric vector of n simulated survival times
#' @export
rtpa <- function(n, eta, phi, alpha, F0, g, QF) {
  u <- runif(n, min = 0, max = 1)
  r <- vector()
  for (i in 1:n) {
    r[i] <- qtpa(u[i], eta, phi, alpha, F0, g, QF)
  }
  return(r)
}
```

## Quick visual check

The chunk below plots the TPA-normal density and hazard function for a
skewed configuration (`alpha != 0.5`), as a sanity check that the functions
above behave as expected.

```{r tpa-demo-plot, fig.cap="TPA-normal density and hazard for eta=1, phi=0.6, alpha=0.35"}
tt <- seq(0.05, 6, length.out = 400)
par(mfrow = c(1, 2))
plot(tt, dtpa(tt, eta = 1, phi = 0.6, alpha = 0.35, f0 = f0_N, g = glog),
     type = "l", xlab = "t", ylab = "density", main = "dtpa()")
plot(tt, htpa(tt, eta = 1, phi = 0.6, alpha = 0.35, f0 = f0_N, F0 = F0_N, g = glog),
     type = "l", xlab = "t", ylab = "hazard", main = "htpa()")
par(mfrow = c(1, 1))
```

\newpage

# Fully Parametric Regression Models

Each function in this section maximises a **negative log-likelihood** built
from `htpa()`/`Htpa()` by a numerical optimiser (`optim()` or `nlminb()`),
returning both the optimiser output and the negative log-likelihood
function itself (useful later for standard errors, see `Conf.Int.FD()`).

All five functions share the same parameter-vector layout for the baseline
distribution: `par[1] = log(eta)`, `par[2] = log(phi)`, `par[3] = logit(alpha)`
(so `eta`, `phi` stay positive and `alpha` stays in (0, 1) during
unconstrained optimisation), followed by regression coefficients.

| Function        | Model | Covariate role(s)                                             |
|------------------|-------|----------------------------------------------------------------|
| `parHH_mle()`   | HH    | `x1` on the time scale, `x2` on the hazard scale (both linear or `x2` profiled out if `psi2.hat` is supplied) |
| `parGH_mle()`   | GH    | same design matrix `x` used on both scales, with separate coefficient vectors |
| `parAFT_mle()`  | AFT   | `x` on the time scale only (pure acceleration)                 |
| `parPH_mle()`   | PH    | `x` on the hazard scale only (pure proportional hazards)       |
| `parAH_mle()`   | AH    | `x` on the time scale, with the hazard-scale effect being the *negative* of the same linear predictor (accelerated hazards) |

```{r par-HH}
#' Maximum likelihood estimation for the parametric Hybrid Hazard (HH) model
#'
#' The HH model allows a covariate to act simultaneously on the time scale
#' (through x1) and the hazard scale (through x2). Semi-parametric use is
#' supported: pass a pre-estimated `psi2.hat` (a numeric vector, one value
#' per observation) to hold the hazard-scale linear predictor fixed while
#' profiling over the remaining parameters -- this is how the code is later
#' combined with the non-parametric estimators in Section 5.
#'
#' @param init numeric vector of starting values for the optimiser
#' @param y observed (possibly censored) survival times
#' @param delta event indicator (1 = event, 0 = censored)
#' @param x1 design matrix for the time-scale covariates (n x d1)
#' @param x2 design matrix for the hazard-scale covariates (n x d2)
#' @param dist baseline distribution: "tpalnorm", "tpallogis" or "tpalLap"
#' @param psi2.hat optional fixed hazard-scale linear predictor (length n);
#'   if supplied, the x2/beta2 block is not estimated
#' @param method optimiser: any `optim()` method, or "nlminb"
#' @param maxit maximum number of iterations (default 10000)
#' @return a list with `OPT` (raw optimiser output) and `nloglik`
#'   (the negative log-likelihood function used, for later reuse e.g. by
#'   `Conf.Int.FD()`)
#' @export
parHH_mle <- function(init, y, delta, x1, x2, dist, psi2.hat = NULL,
                       method = "Nelder-Mead", maxit = 10000, ...) {
  y <- as.vector(y)
  x1 <- as.matrix(x1)
  x2 <- as.matrix(x2)
  d1 <- dim(x1)[2]
  d2 <- dim(x2)[2]

  # --- TPA log-normal HH model --------------------------------------------
  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      # Baseline parameters, transformed back to their natural scale.
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d1)]                 # time-scale regression coefficients
      psi.x1 <- as.vector(x1 %*% beta1)
      exp.psi.x1 <- exp(psi.x1)

      # If psi2.hat is not supplied, estimate beta2 along with everything
      # else; otherwise treat the hazard-scale linear predictor as known
      # (profiled out) -- this is the hook used for semi-parametric fits.
      if (is.null(psi2.hat)) {
        beta2 <- par[(4 + d1):(3 + d1 + d2)]
        psi.x2 <- as.vector(x2 %*% beta2)
      } else {
        psi.x2 <- psi2.hat
      }

      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      # Log-hazard contribution (evaluated at the time-rescaled argument
      # y * exp(psi.x1)) and cumulative-hazard contribution, combined into
      # the negative log-likelihood of a right-censored sample.
      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_N, F0_N,
                    glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_N,
                    glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  # --- TPA log-logistic HH model ------------------------------------------
  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d1)]
      psi.x1 <- as.vector(x1 %*% beta1)
      exp.psi.x1 <- exp(psi.x1)

      if (is.null(psi2.hat)) {
        beta2 <- par[(4 + d1):(3 + d1 + d2)]
        psi.x2 <- as.vector(x2 %*% beta2)
      } else {
        psi.x2 <- psi2.hat
      }

      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_Lo, F0_Lo,
                    glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_Lo,
                    glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  # --- TPA log-Laplace HH model -------------------------------------------
  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d1)]
      psi.x1 <- as.vector(x1 %*% beta1)
      exp.psi.x1 <- exp(psi.x1)

      if (is.null(psi2.hat)) {
        beta2 <- par[(4 + d1):(3 + d1 + d2)]
        psi.x2 <- as.vector(x2 %*% beta2)
      } else {
        psi.x2 <- psi2.hat
      }

      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_La, F0_La,
                    glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_La,
                    glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  # --- Optimisation ---------------------------------------------------------
  if (method != "nlminb") {
    OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  }
  if (method == "nlminb") {
    OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  }
  OUT <- list(OPT = OPT, nloglik = nlog.lik)
  return(OUT)
}
```

```{r par-GH}
#' Maximum likelihood estimation for the parametric General Hazards (GH) model
#'
#' The GH model uses the *same* design matrix `x` for both the time scale
#' and the hazard scale, but with two separate coefficient vectors
#' (`beta1`, `beta2`). This differs from `parHH_mle()`, which allows two
#' different design matrices `x1`/`x2`.
#'
#' @inheritParams parHH_mle
#' @param x common design matrix for both scales (n x d)
#' @export
parGH_mle <- function(init, y, delta, x, dist, method = "Nelder-Mead",
                       maxit = 10000, ...) {
  y <- as.vector(y)
  delta <- as.vector(delta)
  x <- as.matrix(x)
  d <- dim(x)[2]

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d)]          # time-scale coefficients
      beta2 <- par[(4 + d):(3 + 2 * d)] # hazard-scale coefficients

      psi.x1 <- as.vector(x %*% beta1)
      psi.x2 <- as.vector(x %*% beta2)

      exp.psi.x1 <- exp(psi.x1)
      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d)]
      beta2 <- par[(4 + d):(3 + 2 * d)]
      psi.x1 <- as.vector(x %*% beta1)
      psi.x2 <- as.vector(x %*% beta2)

      exp.psi.x1 <- exp(psi.x1)
      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta1 <- par[4:(3 + d)]
      beta2 <- par[(4 + d):(3 + 2 * d)]

      psi.x1 <- as.vector(x %*% beta1)
      psi.x2 <- as.vector(x %*% beta2)

      exp.psi.x1 <- exp(psi.x1)
      exp.psi.dif <- as.vector(exp(-psi.x1 + psi.x2))

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_La, glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- list(OPT = OPT, nloglik = nlog.lik)
  return(OUT)
}
```

```{r par-AFT}
#' Maximum likelihood estimation for the parametric Accelerated Failure
#' Time (AFT) model (time-scale covariate effect only)
#'
#' @inheritParams parGH_mle
#' @param psihat optional fixed extra linear predictor (length n), added to
#'   the estimated x %*% beta term; defaults to a vector of zeros
#' @export
parAFT_mle <- function(init, y, delta, x, dist, method = "Nelder-Mead",
                        psihat = NULL, maxit = 10000, ...) {
  y <- as.vector(y)
  x <- as.matrix(x)
  d <- dim(x)[2]
  if (is.null(psihat)) {
    psi.x2 <- as.matrix(rep(0, length(y)))
  } else {
    psi.x2 <- psihat
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% as.vector(beta))
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      # Note: in the pure AFT model the cumulative hazard is NOT rescaled
      # by exp.psi.x (unlike PH/HH) -- only the argument of the hazard is
      # rescaled by time (y * exp.psi.x). This is what makes it an
      # accelerated-*time* rather than a proportional-*hazard* model.
      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_N, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% as.vector(beta))
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% as.vector(beta))
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_La, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- list(OPT = OPT, nloglik = nlog.lik)
  return(OUT)
}
```

```{r par-PH}
#' Maximum likelihood estimation for the parametric Proportional Hazards
#' (PH) model (hazard-scale covariate effect only)
#'
#' Also usable in a semi-parametric setting: pass a pre-estimated
#' `psihat` to hold the linear predictor fixed.
#'
#' @inheritParams parAFT_mle
#' @export
parPH_mle <- function(init, y, delta, x, dist, psihat = NULL,
                       method = "Nelder-Mead", maxit = 10000, ...) {
  y <- as.vector(y)
  x <- as.matrix(x)
  d <- dim(x)[2]
  if (is.null(psihat)) {
    psi.x2 <- as.matrix(rep(0, length(y)))
  } else {
    psi.x2 <- psihat
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% as.vector(beta))
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      # PH model: time itself is NOT rescaled (unlike AFT); only the
      # hazard is multiplied by exp(psi.x) -- the defining property of a
      # proportional-hazards model.
      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp.psi.x
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      # NOTE (kept as in the original script): unlike the tpalnorm branch
      # above, this branch and the tpalLap branch below check `psihat`
      # again here and re-derive psi.x from it, rather than reusing the
      # psi.x2 computed at the top of the function. Functionally
      # equivalent, but worth knowing if you are diffing against your own
      # modifications.
      if (is.null(psihat)) {
        beta <- par[4:(3 + d)]
        psi.x <- as.vector(x %*% beta)
        exp.psi.x <- exp(psi.x)
      } else {
        psi.x <- psihat
        exp.psi.x <- exp(psihat)
      }

      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp.psi.x
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      if (is.null(psihat)) {
        beta <- par[4:(3 + d)]
        psi.x <- as.vector(x %*% beta)
        exp.psi.x <- exp(psi.x)
      } else {
        psi.x <- psihat
        exp.psi.x <- exp(psihat)
      }

      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_La, glog, log = FALSE) * exp.psi.x
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- list(OPT = OPT, nloglik = nlog.lik)
  return(OUT)
}
```

```{r par-AH}
#' Maximum likelihood estimation for the parametric Accelerated Hazards
#' (AH) model
#'
#' The AH model rescales time by exp(psi.x) (like AFT) AND divides the
#' cumulative hazard contribution by exp(psi.x) (i.e. multiplies by
#' exp(-psi.x)). This combination is what characterises the accelerated
#' hazards model, distinct from AFT and PH.
#'
#' @inheritParams parPH_mle
#' @export
parAH_mle <- function(init, y, delta, x, dist, psihat = NULL,
                       method = "Nelder-Mead", maxit = 10000, ...) {
  y <- as.vector(y)
  x <- as.matrix(x)
  d <- dim(x)[2]
  if (is.null(psihat)) {
    psi.x2 <- as.matrix(rep(0, length(y)))
  } else {
    psi.x2 <- psihat
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% beta)
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE)
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% beta)
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE)
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(par[1]); phi0 <- exp(par[2]); alpha0 <- expit(par[3])
      beta <- par[4:(3 + d)]
      psi.x1 <- as.vector(x %*% beta)
      psi.x <- psi.x1 + psi.x2
      exp.psi.x <- exp(psi.x)

      lhaz0 <- htpa(y * exp.psi.x, eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE)
      chaz0 <- Htpa(y * exp.psi.x, eta0, phi0, alpha0, F0_La, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0, na.rm = TRUE) + sum(chaz0, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- list(OPT = OPT, nloglik = nlog.lik)
  return(OUT)
}
```

\newpage

# Kernel function and semi-/non-parametric estimation

The functions in this section implement **local-likelihood** estimation of
a smooth covariate effect, i.e. the non-parametric building block used when
part of a model's linear predictor is left unspecified. Estimation proceeds
by:

1. picking a **kernel** (weight function) `ker()`,
2. centring the covariate of interest on a grid point `x0` and weighting
   nearby observations more heavily using bandwidth `h`, and
3. locally approximating the unknown smooth function by a second-order
   polynomial (a *local quadratic* fit), whose coefficients are estimated
   by maximising the kernel-weighted log-likelihood.

Each `nonpar*_mle()` function below is the local-likelihood, kernel-weighted
counterpart of the corresponding `par*_mle()` function in Section 4, and
each optionally "profiles out" a parametric part of the covariate effect
(pass its index via `x21.index`/`x11.index` and its fitted coefficients via
`beta2.hat`/`beta1.hat`) -- this is what allows a genuinely **semi-parametric**
model (part linear, part smooth) to be fitted.

```{r kernel-function}
#' Kernel weight functions used for local-likelihood smoothing
#' @param ktype one of "uniform", "Gaussian", "Epan" (Epanechnikov,
#'   the default used throughout), "biweight", "triweight", "cosine"
#' @param x standardised distance (x0 - xi) / h
#' @return kernel weights, same length as x
#' @export
ker <- function(ktype, x) {
  if (ktype == "uniform") {
    result <- dunif(x, -1, 1)
  }
  if (ktype == "Gaussian") {
    result <- dnorm(x)
  } else if (ktype == "Epan") {
    result <- ifelse(abs(x) <= 1, 3 / 4 * (1 - x^2), 0)
  } else if (ktype == "biweight") {
    result <- ifelse(abs(x) <= 1, 15 / 16 * (1 - x^2)^2, 0)
  } else if (ktype == "triweight") {
    result <- ifelse(abs(x) <= 1, 35 / 32 * (1 - x^2)^3, 0)
  } else if (ktype == "cosine") {
    result <- ifelse(abs(x) <= 1, pi / 4 * cos(pi / 2 * x), 0)
  }
  return(result)
}
```

```{r nonpar-HH}
#' Local-likelihood (semi-/non-parametric) estimation of the hazard-scale
#' smooth term in the HH model
#'
#' @param init optional starting values; if NULL, a degree-2 polynomial OLS
#'   fit on the (uncensored) data is used to construct starting values
#' @param y,delta observed times and event indicator
#' @param x1 time-scale design matrix (parametric part, already estimated
#'   via beta1.hat)
#' @param x2 hazard-scale covariate matrix; may combine a parametric block
#'   (columns given by `x21.index`, coefficients `beta2.hat`) and a
#'   non-parametric block (the remaining columns)
#' @param x21.index column indices of x2 that are modelled parametrically;
#'   NULL if the entire hazard-scale effect is non-parametric
#' @param theta.hat fitted (transformed) baseline parameters, i.e.
#'   `c(log(eta), log(phi), logit(alpha))`
#' @param beta1.hat fitted time-scale regression coefficients
#' @param beta2.hat fitted coefficients for the parametric block of x2
#'   (required if x21.index is supplied)
#' @param dist baseline distribution: "tpalnorm" or "tpallogis" (the
#'   log-Laplace case is not implemented for this estimator in the
#'   original code)
#' @param x0 the local grid point at which the smooth function is estimated
#' @param h bandwidth (> 0)
#' @param ktype kernel type, see `ker()`; defaults to "Epan"
#' @param method optimiser, "Nelder-Mead" (default) or "nlminb"
#' @param maxit maximum iterations
#' @return the fitted local polynomial coefficients at x0 (intercept,
#'   linear, quadratic term); the function value at x0 is the first element
#' @export
nonparHH_mle <- function(init = NULL, y, delta, x1, x2, x21.index = NULL,
                          theta.hat, beta1.hat, beta2.hat = NULL, dist,
                          x0, h, ktype = "Epan", method = "Nelder-Mead", maxit = 10000, ...) {
  if (is.na(h) || h <= 0L) stop("invalid 'bandwidth value.' ")
  if (is.null(colnames(x2))) stop("give a column names for the 'covariates'")

  y <- as.vector(y)
  delta <- as.vector(delta)
  x1 <- as.matrix(x1)
  d1 <- dim(x1)[2]
  n <- length(y)

  # Split x2 into a parametric block (x21) and the non-parametric block
  # (x22) that is actually being smoothed at this call.
  if (!is.null(x21.index) | !is.null(beta2.hat)) {
    x21 <- as.matrix(x2[, x21.index])
    d21 <- ncol(x21)
    x22 <- as.matrix(x2[, -c(1:d21)])
    d22 <- ncol(x22)
    psi.x21 <- as.vector(x21 %*% beta2.hat)
  } else {
    psi.x21 <- rep(0, n)
    x22 <- as.matrix(x2)
    d22 <- dim(x22)[2]
  }

  # Centre the non-parametric covariate(s) on the local point x0 and build
  # kernel weights, then a local design matrix up to the quadratic term.
  newx0 <- x22 - t(matrix(rep(x0, d22 * n), d22, n))
  ker.h <- as.vector(apply(newx0 / h, MARGIN = 2, FUN = ker, ktype = ktype) / h)
  newx <- cbind(1, newx0, newx0^2) # local quadratic (up to second derivative)

  if (is.null(init)) {
    fit.poly <- lm(y[delta == 1] ~ poly(newx0[delta == 1], degree = 2))
    init <- as.numeric(fit.poly$coefficients)
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b2 <- par
      psi.x22 <- as.vector(newx %*% b2)

      psi.x2 <- psi.x21 + psi.x22
      psi.x1 <- as.vector(x1 %*% beta1.hat)
      exp.psi.x1 <- exp(psi.x1)
      exp.psi.dif <- exp(-psi.x1 + psi.x2)

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      # Kernel-weighted (local) negative log-likelihood.
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b2 <- par
      psi.x22 <- as.vector(newx %*% b2)

      psi.x2 <- psi.x21 + psi.x22
      psi.x1 <- as.vector(x1 %*% beta1.hat)
      exp.psi.x1 <- exp(psi.x1)
      exp.psi.dif <- exp(-psi.x1 + psi.x2)

      lhaz0 <- htpa(y * exp.psi.x1, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y * exp.psi.x1, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp.psi.dif
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- OPT$par
  return(OUT)
}
```

```{r nonpar-PH}
#' Local-likelihood estimation of the hazard-scale smooth term in the PH
#' model (no time-scale rescaling)
#' @inheritParams nonparHH_mle
#' @param x2 as in nonparHH_mle
#' @export
nonparPH_mle <- function(init = NULL, y, delta, x2, x21.index = NULL,
                          theta.hat, beta2.hat = NULL, dist,
                          x0, h, ktype = "Epan", method = "Nelder-Mead", maxit = 10000, ...) {
  if (is.na(h) || h <= 0L) stop("invalid 'bandwidth value.' ")
  if (is.null(colnames(x2))) stop("give a column names for the 'covariates'")

  y <- as.vector(y)
  delta <- as.vector(delta)
  n <- length(y)

  if (!is.null(x21.index) | !is.null(beta2.hat)) {
    x21 <- as.matrix(x2[, x21.index])
    d21 <- ncol(x21)
    x22 <- as.matrix(x2[, -c(1:d21)])
    d22 <- ncol(x22)
    psi.x21 <- as.vector(x21 %*% beta2.hat)
  } else {
    psi.x21 <- rep(0, n)
    x22 <- as.matrix(x2)
    d22 <- dim(x22)[2]
  }

  newx0 <- x22 - t(matrix(rep(x0, d22 * n), d22, n))
  ker.h <- as.vector(apply(newx0 / h, MARGIN = 2, FUN = ker, ktype = ktype) / h)
  newx <- cbind(1, newx0, newx0^2)

  if (is.null(init)) {
    fit.poly <- lm(y[delta == 1] ~ poly(newx0[delta == 1], degree = 2))
    init <- as.numeric(fit.poly$coefficients)
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b2 <- par
      psi.x22 <- as.vector(newx %*% b2)
      psi.x2 <- psi.x21 + psi.x22

      # PH model: time is not rescaled; the hazard is scaled multiplicatively.
      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp(psi.x2)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b2 <- par
      psi.x22 <- as.vector(newx %*% b2)
      psi.x2 <- psi.x21 + psi.x22

      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp(psi.x2)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b2 <- par
      psi.x22 <- as.vector(newx %*% b2)
      psi.x2 <- psi.x21 + psi.x22

      lhaz0 <- htpa(y, eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE) + psi.x2
      chaz0 <- Htpa(y, eta0, phi0, alpha0, F0_La, glog, log = FALSE) * exp(psi.x2)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- OPT$par
  return(OUT)
}
```

```{r nonpar-AFT}
#' Local-likelihood estimation of the time-scale smooth term in the AFT
#' model
#' @inheritParams nonparHH_mle
#' @param x design matrix; may combine a parametric block (`x11.index`,
#'   `beta1.hat`) and a non-parametric block
#' @export
nonparAFT_mle <- function(init = NULL, y, delta, x, x11.index = NULL,
                           theta.hat, beta1.hat = NULL, dist,
                           x0, h, ktype = "Epan", method = "Nelder-Mead", maxit = 10000, ...) {
  if (is.na(h) || h <= 0L) stop("invalid 'bandwidth value.' ")
  if (is.null(colnames(x))) stop("give a column names for the 'covariates'")

  y <- as.vector(y)
  delta <- as.vector(delta)
  n <- length(y)

  if (!is.null(x11.index) | !is.null(beta1.hat)) {
    x11 <- as.matrix(x[, x11.index])
    d11 <- ncol(x11)
    x12 <- as.matrix(x[, -c(1:d11)])
    d12 <- ncol(x12)
    psi.x11 <- as.vector(x11 %*% beta1.hat)
  } else {
    psi.x11 <- rep(0, n)
    x12 <- as.matrix(x)
    d12 <- dim(x12)[2]
  }

  newx0 <- x12 - t(matrix(rep(x0, d12 * n), d12, n))
  ker.h <- as.vector(apply(newx0 / h, MARGIN = 2, FUN = ker, ktype = ktype) / h)
  newx <- cbind(1, newx0, newx0^2)

  if (is.null(init)) {
    fit.poly <- lm(y[delta == 1] ~ poly(newx0[delta == 1], degree = 2))
    init <- as.numeric(fit.poly$coefficients)
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_N, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_Lo, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE) + psi.x
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_La, glog, log = FALSE)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- OPT$par
  return(OUT)
}
```

```{r nonpar-AH}
#' Local-likelihood estimation of the time-scale smooth term in the AH
#' model
#' @inheritParams nonparAFT_mle
#' @export
nonparAH_mle <- function(init = NULL, y, delta, x, x11.index = NULL,
                          theta.hat, beta1.hat = NULL, dist,
                          x0, h, ktype = "Epan", method = "Nelder-Mead", maxit = 10000, ...) {
  if (is.na(h) || h <= 0L) stop("invalid 'bandwidth value.' ")
  if (is.null(colnames(x))) stop("give a column names for the 'covariates'")

  y <- as.vector(y)
  delta <- as.vector(delta)
  n <- length(y)

  if (!is.null(x11.index) | !is.null(beta1.hat)) {
    x11 <- as.matrix(x[, x11.index])
    d11 <- ncol(x11)
    x12 <- as.matrix(x[, -c(1:d11)])
    d12 <- ncol(x12)
    psi.x11 <- as.vector(x11 %*% beta1.hat)
  } else {
    psi.x11 <- rep(0, n)
    x12 <- as.matrix(x)
    d12 <- dim(x12)[2]
  }

  newx0 <- x12 - t(matrix(rep(x0, d12 * n), d12, n))
  ker.h <- as.vector(apply(newx0 / h, MARGIN = 2, FUN = ker, ktype = ktype) / h)
  newx <- cbind(1, newx0, newx0^2)

  if (is.null(init)) {
    fit.poly <- lm(y[delta == 1] ~ poly(newx0[delta == 1], degree = 2))
    init <- as.numeric(fit.poly$coefficients)
  }

  if (dist == "tpalnorm") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      # AH model: time rescaled by exp(psi.x), cumulative hazard divided
      # by exp(psi.x) (i.e. multiplied by exp(-psi.x)).
      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_N, F0_N, glog, log = TRUE)
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_N, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpallogis") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_Lo, F0_Lo, glog, log = TRUE)
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_Lo, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (dist == "tpalLap") {
    nlog.lik <- function(par) {
      if (anyNA(par)) return(Inf)
      eta0 <- exp(theta.hat[1]); phi0 <- exp(theta.hat[2]); alpha0 <- expit(theta.hat[3])
      b1 <- par
      psi.x12 <- as.vector(newx %*% b1)
      psi.x <- psi.x11 + psi.x12

      lhaz0 <- htpa(y * exp(psi.x), eta0, phi0, alpha0, f0_La, F0_La, glog, log = TRUE)
      chaz0 <- Htpa(y * exp(psi.x), eta0, phi0, alpha0, F0_La, glog, log = FALSE) * exp(-psi.x)
      lhaz0[is.infinite(lhaz0)] <- NA
      chaz0[is.infinite(chaz0)] <- NA
      val <- -sum(delta * lhaz0 * ker.h, na.rm = TRUE) + sum(chaz0 * ker.h, na.rm = TRUE)
      return(val)
    }
  }

  if (method != "nlminb") OPT <- optim(init, nlog.lik, control = list(maxit = maxit), method = method)
  if (method == "nlminb") OPT <- nlminb(init, nlog.lik, control = list(iter.max = maxit))
  OUT <- OPT$par
  return(OUT)
}
```

\newpage

# Post-estimation tools

```{r ase}
#' Average squared error between two survival curves
#' @param true_sp true/reference survival probabilities
#' @param pred_sp predicted survival probabilities
#' @return a single numeric value (the mean squared difference)
#' @export
surv_ase <- function(true_sp, pred_sp) {
  if (length(true_sp) != length(pred_sp))
    stop(" Error: surv_sp and pred_sp must have equal length")
  d <- (true_sp - pred_sp)^2
  return(ase = mean(d, na.rm = TRUE))
}
```

```{r predict-fun}
#' Predicted survival, hazard, cumulative hazard, or quantile from a fitted
#' TPA (or "EW") model
#'
#' @param tau probability(-ies) for `type = "quantile"` (in (0,1))
#' @param theta transformed baseline parameters `c(log(eta), log(phi),
#'   logit(alpha))` -- NOTE: the function body actually reads these from an
#'   object called `par`, not from the `theta` argument (see the "Known
#'   limitations" note in Section 9); supply a global/enclosing `par`
#'   vector with the natural-scale baseline parameters `c(eta, phi, alpha)`
#'   until this is corrected
#' @param time vector of times at which to predict
#' @param psihat.1 fitted time-scale linear predictor (vector, length n)
#' @param psihat.2 fitted hazard-scale linear predictor (vector, length n)
#' @param dist baseline distribution: "tpalnorm", "tpallogis", or "EW"
#'   (Exponentiated Weibull -- requires a package providing `pexpweibull`/
#'   `hexpweibull`/`chexpweibull`/`qexpweibull`, which is NOT loaded
#'   anywhere in this file; see Section 9)
#' @param type one of "survival" (default), "hazard", "cmhazard", "quantile"
#' @return numeric vector of predictions, one per (sorted) time / tau value
#' @export
predict_fun <- function(tau = NULL, theta, time, psihat.1, psihat.2,
                         dist, type = "survival", ...) {
  time <- sort(time)
  psihat.dif <- psihat.2 - psihat.1

  if (type == "survival") {
    if (dist == "tpalnorm") {
      value <- ptpa(time * exp(psihat.1), par[1], par[2], par[3],
                    F0_N, glog, lower.tail = FALSE)^exp(psihat.dif)
    } else if (dist == "tpallogis") {
      value <- ptpa(time * exp(psihat.1), par[1], par[2], par[3],
                    F0_Lo, glog, lower.tail = FALSE)^exp(psihat.dif)
    } else if (dist == "EW") {
      value <- 1 - pexpweibull(time * exp(psihat.1), par[1], par[2], par[3])^exp(psihat.dif)
    } else stop("Baseline distribution is not provided")
  } else if (type == "hazard") {
    if (dist == "tpalnorm") {
      value <- htpa(time * exp(psihat.1), par[1], par[2], par[3],
                    f0_N, F0_N, glog) * exp(psihat.2)
    } else if (dist == "tpallogis") {
      value <- htpa(time * exp(psihat.1), par[1], par[2], par[3],
                    f0_Lo, F0_Lo, glog) * exp(psihat.2)
    } else if (dist == "EW") {
      value <- hexpweibull(time * exp(psihat.1), par[1], par[2], par[3]) * exp(psihat.2)
    } else stop("Baseline distribution is not provided")
  } else if (type == "cmhazard") {
    if (dist == "tpalnorm") {
      value <- Htpa(time * exp(psihat.1), par[1], par[2], par[3], F0_N, glog) * exp(psihat.dif)
    } else if (dist == "tpallogis") {
      value <- Htpa(time * exp(psihat.1), par[1], par[2], par[3], F0_Lo, glog) * exp(psihat.dif)
    } else if (dist == "EW") {
      value <- chexpweibull(time * exp(psihat.1), par[1], par[2], par[3]) * exp(psihat.dif)
    } else stop("Baseline distribution is not provided")
  } else if (type == "quantile") {
    if (dist == "tpalnorm") {
      value <- exp(-psihat.1) * qtpa(1 - exp(log(1 - tau) / exp(psihat.dif)), par[1], par[2], par[3],
                                      F0_N, glog, glog.inv, QF_N)
    } else if (dist == "tpallogis") {
      value <- exp(-psihat.1) * qtpa(1 - exp(log(1 - tau) / exp(psihat.dif)), par[1], par[2], par[3],
                                      F0_Lo, glog, glog.inv, QF_Lo)
    } else if (dist == "EW") {
      value <- exp(-psihat.1) * qexpweibull(1 - exp(log(1 - tau) / exp(psihat.dif)), par[1], par[2], par[3])
    } else stop("Baseline distribution is not provided")
  }
  return(value)
}
```

```{r predict-plot}
#' Simple scatter/line plot of a predicted quantity against time
#' @param object predicted values (e.g. output of predict_fun)
#' @param time the corresponding times
#' @param xlab,ylab axis labels
#' @export
predict_plot <- function(object, time, xlab = "Time", ylab = "Survival probability") {
  ord <- order(time)
  plot(time[ord], object[ord], xlab = xlab, ylab = ylab)
}
```

```{r trapezoid}
#' Numerical integration of an estimated derivative curve via the
#' trapezoidal rule, to recover psi(x) from an estimate of psi'(x)
#'
#' Two implementations are kept, matching the original script:
#' `trap_fun()` uses `pracma::trapz()`, `trap_phihat()` is a
#' dependency-free hand-written equivalent that additionally requires x0
#' to already be sorted.
#'
#' @param x0 vector of grid values (the points at which the derivative was
#'   estimated)
#' @param dpsi.hat estimated derivative of psi at each x0
#' @return cumulative (trapezoidal) integral of dpsi.hat from the smallest
#'   x0 up to each grid point, i.e. an estimate of psi(x0)
#' @export
trap_fun <- function(x0, dpsi.hat) {
  require(pracma)
  if (is.unsorted(x0)) {
    ord <- order(x0)
    x0 <- x0[ord]
    dpsi.hat <- dpsi.hat[ord]
  }

  m <- length(x0)
  psi.hat <- as.null(x0)
  psi.hat[1] <- (x0[1] / 2) * dpsi.hat[1]
  for (k in 2:m) {
    s.x <- sapply(2:k, function(j) trapz(x = x0[(j - 1):j], y = dpsi.hat[(j - 1):j]))
    psi.hat[k] <- sum(s.x)
  }
  return(psi.hat = as.numeric(psi.hat))
}

#' @rdname trap_fun
#' @export
trap_phihat <- function(x0, dpsi.hat) {
  if (!is.sorted(x0))
    stop(" 'x0' should be sorted")
  m <- length(x0)
  psi.hat <- as.null(x0)
  psi.hat[1] <- (x0[1] / 2) * dpsi.hat[1]
  for (k in 2:m) {
    s.x <- sapply(2:k, function(j) (x0[j] - x0[j - 1]) * (dpsi.hat[j] + dpsi.hat[j - 1]) / 2)
    psi.hat[k] <- sum(s.x)
  }
  return(psi.hat = as.numeric(psi.hat))
}
```

```{r conf-int}
#' Wald-type confidence intervals via a finite-difference Hessian
#'
#' Computes the observed Fisher information as the inverse of a
#' numerically approximated Hessian of the negative log-likelihood
#' (via `nlme::fdHess()`), then builds symmetric Wald confidence intervals
#' around the MLE.
#'
#' @param nlogLik the negative log-likelihood function (e.g. the
#'   `nloglik` element returned by any `par*_mle()` function)
#' @param mle the maximum-likelihood estimate (vector), i.e. `OPT$par` (or
#'   `OPT$OPT$par` from the par*_mle() output list)
#' @param level confidence level, default 0.95
#' @return a matrix with columns Estimates, Std. Error, Lower, Upper; one
#'   row per parameter
#' @export
Conf.Int.FD <- function(nlogLik, mle, level = 0.95, ...) {
  require(nlme)
  sd.int <- abs(qnorm(0.5 * (1 - level)))
  HESS <- fdHess(pars = mle, fun = nlogLik)
  Fisher.Info <- solve(HESS$Hessian)
  se <- sqrt(diag(Fisher.Info))
  U <- mle + sd.int * se
  L <- mle - sd.int * se
  C.I <- cbind(mle, se, L, U)
  names.row <- paste0("par", seq_along(1:length(mle)))
  rownames(C.I) <- names.row
  colnames(C.I) <- c("Estimates", "Std. Error", "Lower", "Upper")
  return(C.I)
}
```

\newpage

# Worked example

This example simulates right-censored survival data from a **TPA-normal
Proportional Hazards** model with a single binary covariate, fits the model
with `parPH_mle()`, obtains standard errors and 95% confidence intervals
with `Conf.Int.FD()`, and predicts/plots the fitted survival curve for each
group using `htpa()`/`ptpa()` directly (a thin, corrected stand-in for
`predict_fun()` -- see the note on `predict_fun()` in Section 9).

```{r worked-example-simulate}
set.seed(2026)
n <- 300

# One binary covariate (e.g. treatment arm).
x <- matrix(rbinom(n, 1, 0.5), ncol = 1)
colnames(x) <- "trt"

# True (data-generating) parameters: baseline TPA-normal parameters and a
# single hazard-scale (PH) log-hazard-ratio.
eta_true   <- 1.2
phi_true   <- 0.6
alpha_true <- 0.4
beta_true  <- -0.7   # trt = 1 halves the hazard, roughly

# Simulate event times: the PH construction used in parPH_mle() rescales
# only the hazard, not time, so we invert the model's own survival
# function group-by-group using the quantile function qtpa().
u <- runif(n)
psi.x <- as.vector(x %*% beta_true)
# Under this PH construction, S(t | x) = S0(t)^exp(psi.x), so inverting:
# u = S0(t)^exp(psi.x)  =>  S0(t) = u^(1/exp(psi.x))  =>  t = QF0(1 - S0(t))
target_tau <- 1 - u^(1 / exp(psi.x))
true_time <- sapply(target_tau, function(p) {
  qtpa(p, eta_true, phi_true, alpha_true, F0_N, glog, glog.inv, QF_N)
})

# Independent random censoring.
cens_time <- rexp(n, rate = 0.15)
y <- pmin(true_time, cens_time)
delta <- as.numeric(true_time <= cens_time)

cat("Simulated n =", n, " | events =", sum(delta), " | censored =", sum(1 - delta), "\n")
```

```{r worked-example-fit}
# Starting values: log(eta), log(phi), logit(alpha), beta.
init <- c(log(1), log(1), 0, 0)

fit <- parPH_mle(init = init, y = y, delta = delta, x = x,
                  dist = "tpalnorm", method = "nlminb", maxit = 5000)

est <- fit$OPT$par
names(est) <- c("log(eta)", "log(phi)", "logit(alpha)", "beta_trt")
knitr::kable(round(cbind(Estimate = est), 4),
             caption = "MLE on the transformed (unconstrained) scale")

# Back-transform to the natural scale for interpretation.
natural <- c(eta = exp(est[1]), phi = exp(est[2]), alpha = expit(est[3]),
             hazard_ratio_trt = exp(est[4]))
knitr::kable(round(cbind(Estimate = natural), 4),
             caption = "Natural-scale parameters (hazard_ratio_trt is exp(beta_trt))")
```

```{r worked-example-ci}
ci <- Conf.Int.FD(nlogLik = fit$nloglik, mle = fit$OPT$par)
rownames(ci) <- c("log(eta)", "log(phi)", "logit(alpha)", "beta_trt")
knitr::kable(round(ci, 4),
             caption = "Wald 95% confidence intervals (transformed scale)")
```

```{r worked-example-predict, fig.cap="Fitted TPA-normal PH survival curves by treatment group"}
eta_hat   <- exp(est[1]); phi_hat <- exp(est[2]); alpha_hat <- expit(est[3])
beta_hat  <- est[4]

tt <- seq(0.01, quantile(y, 0.95), length.out = 200)

# PH model: S(t | x) = S0(t)^exp(x * beta_hat)
S0 <- ptpa(tt, eta_hat, phi_hat, alpha_hat, F0_N, glog, lower.tail = FALSE)
S_trt0 <- S0^exp(0 * beta_hat)  # trt = 0
S_trt1 <- S0^exp(1 * beta_hat)  # trt = 1

plot(tt, S_trt0, type = "l", lwd = 2, ylim = c(0, 1),
     xlab = "Time", ylab = "Survival probability",
     main = "Fitted survival curves: TPA-normal PH model")
lines(tt, S_trt1, lwd = 2, lty = 2, col = "steelblue")
legend("topright", legend = c("trt = 0", "trt = 1"),
       lwd = 2, lty = c(1, 2), col = c("black", "steelblue"), bty = "n")
```

The fitted `beta_trt` should recover a value in the neighbourhood of the
true `r beta_true` used to simulate the data, subject to sampling
variability at `n = 300`; the printed 95% confidence interval for
`beta_trt` should typically cover `r beta_true`.

\newpage

# Function reference table

```{r ref-table, echo=FALSE}
ref <- data.frame(
  Function = c("f0_N/F0_N/QF_N", "f0_La/F0_La/QF_La", "f0_Lo/F0_Lo/QF_Lo",
               "glog/glog.inv, glogit/glogit.inv", "dtpa", "ptpa", "htpa", "Htpa",
               "qtpa", "rtpa", "parHH_mle", "parGH_mle", "parAFT_mle", "parPH_mle",
               "parAH_mle", "ker", "nonparHH_mle", "nonparPH_mle", "nonparAFT_mle",
               "nonparAH_mle", "surv_ase", "predict_fun", "predict_plot",
               "trap_fun / trap_phihat", "Conf.Int.FD"),
  Purpose = c(
    "Normal reference density/CDF/quantile",
    "Laplace reference density/CDF/quantile",
    "Logistic reference density/CDF/quantile",
    "Link functions for the time scale",
    "TPA family density",
    "TPA family CDF / survival function",
    "TPA family hazard function",
    "TPA family cumulative hazard function",
    "TPA family quantile function",
    "Random draws from the TPA family",
    "Parametric MLE, Hybrid Hazard model",
    "Parametric MLE, General Hazards model",
    "Parametric MLE, Accelerated Failure Time model",
    "Parametric MLE, Proportional Hazards model",
    "Parametric MLE, Accelerated Hazards model",
    "Kernel weight functions",
    "Local-likelihood MLE, HH model (hazard-scale smooth term)",
    "Local-likelihood MLE, PH model (hazard-scale smooth term)",
    "Local-likelihood MLE, AFT model (time-scale smooth term)",
    "Local-likelihood MLE, AH model (time-scale smooth term)",
    "Average squared error between two survival curves",
    "Predicted survival/hazard/cum.hazard/quantile (see Sec. 9 note)",
    "Simple plot of predicted quantity vs. time",
    "Trapezoidal-rule numerical integration of an estimated derivative",
    "Finite-difference Wald confidence intervals"
  )
)
knitr::kable(ref, caption = "One-line summary of every function in this file")
```

\newpage

# Notes on the original code

Nothing in the executable code above was modified relative to the original script `Topic3_sourcecode.R` (only comments, spacing, and file organisation changed); the two items below are pointed out so you can decide whether to work around them or patch them yourself.

1. **`predict_fun()` reads baseline parameters from a variable called
   `par`, not from its own `theta` argument.** In the original script the
   function signature includes a `theta` parameter that is documented but
   never used inside the function body; instead, every branch references
   an object literally called `par`. This works only if a suitably named
   vector `par` (natural-scale `c(eta, phi, alpha)`) happens to exist in
   the calling environment (for example, left over from a prior `optim()`
   call that used `par` as its variable name). In the worked example above
   we avoided `predict_fun()` for this reason and instead called
   `ptpa()`/`htpa()`/`Htpa()` directly with the fitted parameters. If you
   want to use `predict_fun()` as-is, define a variable named exactly
   `par <- c(eta_hat, phi_hat, alpha_hat)` before calling it.

2. **The `dist = "EW"` branches of `predict_fun()` call functions that are
   not defined or imported anywhere in this file** (`pexpweibull`,
   `hexpweibull`, `chexpweibull`, `qexpweibull` — these come from an
   Exponentiated-Weibull package that the original script never loads).
   These branches will fail with an "object not found" error unless you
   load a package that provides those four functions yourself.

3. **`nonparHH_mle()` and `nonparPH_mle()` implement `dist = "tpalnorm"`
   and `"tpallogis"` (and, for `nonparPH_mle()`, `"tpalLap"`)**, but do not
   implement a TPA log-Laplace branch for `nonparHH_mle()`. Calling
   `nonparHH_mle(..., dist = "tpalLap")` will error because `nlog.lik` is
   never assigned. This mirrors the original script exactly.

4. **`trap_phihat()` calls `is.sorted()`**, which is not a base-R function
   (base R provides `is.unsorted()`, whose negation is what's presumably
   intended). This function will error immediately if called, in both the
   original script and this rewrite. Prefer `trap_fun()` (which uses
   `pracma::trapz()` and handles unsorted input automatically) unless you
   first fix this line yourself.

5. **All optimisers use `method = "Nelder-Mead"` by default**, a
   derivative-free method that is robust to a poorly chosen starting point
   but can be slow to converge for larger parameter vectors; for
   production use, consider trying `method = "nlminb"` or `method =
   "BFGS"` as well and comparing results, especially once you add more
   than a handful of covariates.

# Original repository

* Source: <https://github.com/Ewnetu-github/tpaHH>
* Original single file: `Topic3_sourcecode.R`
* Original README: *"This is a source code for a hybrid hazard-based models
  using two-piece distributions. We proposed parametric, semi-parametric
  and non-parametric regression functions for the time-scale and relative
  hazard-scale change. Three parametric baseline distributions with a
  log-link function are included in the code: TPA normal, TPA logistic and
  TPA Laplace distributions."*

This document (`tpaHH_two_piece_hazard_models.Rmd`) can be re-knitted at
any time with:

```r
rmarkdown::render("tpaHH_two_piece_hazard_models.Rmd", output_format = "pdf_document")
rmarkdown::render("tpaHH_two_piece_hazard_models.Rmd", output_format = "word_document")
```
